Computational methods in landscape ecology

### 2024-09-18 Sections/01-introduction.qmd
Some cleaning intro and patterns

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
<!-- MH: This sentence needs some editing to be more concise-->
<!-- JN: the examples in the following section should be ordered in a similar way as the topics in the paper -->
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, pattern-process links, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing these developments 
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2002KeyIssues; @Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them. 
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.

### 2024-09-18 Sections/02-spatial-patterns.qmd
Some cleaning intro and patterns

<!-- JN: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
<!-- MH: I think we decided to not reference software here but in a separate section -->
<!-- MH: Alternatively, if we move the software section to the beginning, we could reference software here -->

@@

Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
However, spatial patterns are not only a result of processes, but could also be a driver of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989LandscapeEcology].

## Landscape metrics
<!-- Hesselbarth -->

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural world [@Frazier2017LandscapeMetrics].

Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].

## Entropy
<!-- Nowosad, Bosch -->

<!-- JN: ask T. Stepinski for review of this section later -->

@@

<!--JN: I would suggest to include the following paragraph, which I think is important to show the current state of the affairs.-->
The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@Stepinski2022CurbYour and @Huckeba2023MultiscaleSpatial showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

<!-- Kolmogorov, etc. -->
<!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->

@@

Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990SpatialTemporal; @Wickham1994MappingAnalyzing; @Jasiewicz2015GeoPATToolbox; @Remmel2020DistributionsHyperlocal; @Nowosad2021MotifOpensource].
These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between these signatures.

Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].

R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
The *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.

<!--JN: I would restructure the above paragraph to highlight what can and cannot be done with the current software. You may decided to replace the above paragraph with the following one:-->
<!--
R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], *rasterio* [@gillies2019] in Python, and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia. 
-->
<!--raster package in Julia??-->
<!--
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations, such as spatial clustering, spatial regression, or spatial interpolation.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
-->

## Landscape mosaic
<!-- Bosch -->

<!-- JN: this section seems a bit disconnected from the rest of the text. One option may be to integrate it a bit with the landscape metrics section, and just highlight the most important points.-->

<!-- MH: Is the landscape mosaic after e.g., Vogt et al. 10.1371/journal.pone.0304215 the same as the patch mosaic model and landscape metrics? I think there is some confusion here. We already discuss metrics above. I think it would be more valuable to discuss the LLM and the most recent additions after Vogt et al. here? -->

<!-- https://doi.org/10.1371/journal.pone.0304215 -->
Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998QuantifyingLandscape; @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology; @Lausch2015UnderstandingQuantifying; @Frazier2017LandscapeMetrics].
<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation. Let us go through them in a different order: -->
Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009SpatialMetrics].

### Sensitivity to spatial scale and thematic resolution

@@

### Redundancy
<!-- (iii) selecting metrics -->

<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->

<!-- TODO

- Factor analysis of Riitters et al. [@Riitters1995] found six main components: average patch compaction, image texture, average patch shape, patch perimeter-area scaling, number of attribute classes and large-patch density-area scaling.

- Cushman et al. [@Cushman2008Parsimony] not only isolate the components but also assess their strength (proportion of variance explained), universality (how present they are accross landscapes) and consistency (how stable they are accross landscapes). At the class level, they find seven "highly universal and consistent structure components" (page 700), namely edge contrast, patch shape complexity, aggregation, nearest neighbor distance, patch dispersion, large patch dominance and neighborhood similarity. At the landscape level, five components are identified, namely contagion/diversity (similar to image texture in Riitters et al. [@Riitters1995], large patch dominance, interspersion/juxtaposition, edge contrast, patch shape variability, proximity and nearest neighbor distance. Overall, components are more consistent and universal at the landscape level than at the class level.

- Uuemaa et al. [@Uuemaa2011] find four main components: dominance (also identified by Riitters et al. [@Riitters1995] and Cushman et al. [@Cushman2008Parsimony]), contrast, shape complexity and composition.

- Amount (composition) and adjacency (configuration) are the fundamental elements of pattern [@Riitters2019]

- Entropy [@Nowosad2019a]

- Spatial signatures [@Nowosad2021a]

-->

### Interpretability
<!-- (iv) interpretatability -->

@@

## Surface metrics
<!-- Bosch, Hesselbarth -->

<!-- JN: I would suggest to move all of the metrics subsections (landscape, entropy, surface, vector) to the beginning of the section, and just then do the operations on spatial patterns. This would make the text more coherent. -->

<!-- MH: What about Spectral and Wavelet Analysis? -->

## Vector metrics
<!-- Nowosad, Hesselbarth -->

>>>>

@@

Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
However, spatial patterns are not only a result of processes, but could also be drivers of them.
Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].

## Landscape metrics

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
<!-- Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource]. -->
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified regarding their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

<!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995], principal component analysis [@Cushman2008Parsimony], the variance inflation factor [@Plexida2014], or classification trees [@Breiman2017] were used to identify main components of landscape patterns.
<!-- MH: Add one sentence summarising studies above if possible -->
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->

Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->

<!-- Short section about Landscape mosaic here based on https://doi.org/10.1371/journal.pone.0304215 -->

## Entropy

<!-- JN: ask T. Stepinski for review of this section later -->
<!-- MH: Maybe we need to shorten this a bit -->

@@

The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@Stepinski2022CurbYour and @Huckeba2023MultiscaleSpatial showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.

Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].

@@

## Surface metrics

## Vector metrics

@@

Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
These operations are based on spatial signatures which are multi-numerical representations of landscape pattern and dissimilarity measures which are functions that quantify differences between the signatures.
Spatial patterns can be compared between different areas, or for the same area at two different moments in time.

Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@gillies2019] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia. 
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

## Sensitivity to spatial scale and thematic resolution

@@

## Interpretability
<!-- (iv) interpretatability -->

@@
